Cover Times of Random Walks

### 2019-04-21 paper.tex
figures

\subsection{Matthews Method on Binary Trees}
\todo{graph example of binary tree}

\cite{LP17}
Consider a complete unweighted binary tree with a single root $\rho$.
Each vertex has two children if it is distance less than $k$
from the root and no children if it is a leaf (at distance $k$ from the root).
There are $n = 1 + 2 + 4 + ... + 2^k = 2^{k+1} - 1$ vertices.

@@

\todo{graph example of spanning tree}

A spanning tree $\T$ is a subset of a graph with all edges
in $V(G)$ connected by the minimum number of edges in $E(G)$.
For the connected graphs we consider, the minimum number of edges
needed to connect all vertices is $n-1$.
Formally, $V(\T) = V(G)$ and $E(\T) \subseteq E(G)$
such that $|E(\T)| = n-1$ and $\T$ is connected.

@@

The expected time of this particular path on the original graph is certainly less than
or equal to the cover time. Thus
\begin{align}
\tcov &\leq \sum_{i=0}^{2n-3} \E_{v_i}[T_{v_{i+1}}]. \nonumber
\end{align}

@@

Using \cref{eqn:commute}, $E_x[T_y] + \E_y[T_x] = C \Rxy$,
\begin{align}
\tcov &\leq C \sum_{x,y \in E(\T)} \Rxy. \nonumber
\end{align}

@@

If the graph is unweighted, then $C_{x,y} = 1$ and $C = 2 |E(G)|$.
Thus
\begin{align}
\tcov \leq 2 |E(G)| (n-1). \nonumber
\end{align}

@@

\section{Distributional Aspects}
\cite{Du11}
Consider a random walk on the complete graph with self-edges.
Call $r$ the number of steps since we began the random walk.
Let $A_i$ be the event that vertex $i$ is not visited and
$N_n$ the number of unvisited vertices.
Then
\begin{align}
P(A_i) = (1-1/n)^r \nonumber
\end{align}
and
\begin{align}
\E[N_n] = n(1-1/n)^r. \nonumber
\end{align}

If $r/n \rightarrow c$, then
\begin{align}
\E[N_n]/n &= (1-1/n)^r = [(1-1/n)^n]^{r/n} \rightarrow (e^{-1})^c \nonumber \\
\E[N_n]/n &\rightarrow e^{-c}. \nonumber
\end{align}

We compute the variance of $N_n$.
Let $\I_m$ be the indicator random variable for $A_m$.
First find $E[N_n^2]$,
\begin{align}
\E[N_n^2] &= \E \left[ \left(\sum_{m=1}^n \I_m \right) ^2 \right] =
\E \left[ \sum_{m=1}^n \sum_{k=1}^n \I_m \I_k \right] =
\sum_{1 \leq k, m \leq n} \pr(A_k \cap A_m). \nonumber
\end{align}

Then
\begin{align}
\Var[N_n] &= \E[N_n^2] - \E[N_n]^2 \nonumber \\
&= \sum_{1 \leq k, m \leq n} \pr(A_k \cap A_m) - P(A_k)P(A_m) \nonumber \\
&= n(n-1) \{ \pr(A_k \cap A_m) - P(A_k)P(A_m) \} + 
n \{ \pr(A_k \cap A_m) - P(A_k)P(A_m) \} \nonumber \\
&= n(n-1) \{ (1-2/n)^r - (1-1/n)^{2r} \} + 
n \{ (1-1/n)^r  - (1-1/n)^{2r} \}. \nonumber
\end{align}
where $k\neq m$ in the first term and $k=m$ in the second.

@@

Thus we conclude $N_n/n \rightarrow \E[N_n/n] = e^{-c}$ in probability.

The Poisson approximation of the binomial implies that if $n \rightarrow \infty$
and $r/n \rightarrow c$, then the number of visits to each vertex will approach
a Poisson distribution with mean $c$.
Then it intuitively follows that the proportion of visits to each vertex
approaches $e^{-c}$.

\begin{theorem}\label{thm:pois}
If $ne^{-r/n} \rightarrow \lambda \in [0, \infty)$, then the number of unvisited
vertices approaches a Poisson distribution with mean $\lambda$.
\end{theorem}

\begin{proof}
Observe that the probability that there are $k$ unvisited vertices
\begin{align}
(1-k/n)^r \nonumber.
\end{align}

@@

We want to show that 
\begin{align}
{n \choose m} \left(1-\frac{m}{n}\right)^r \rightarrow \lambda ^m /m! \nonumber
\end{align}

We begin with the more straightforward upper bound.
Notice that $(1-x) \leq e^{-x}$ and recall our assumption that
$ne^{r/n} \rightarrow \lambda$.
Then
\begin{align}
{n \choose m} \left(1-\frac{m}{n}\right)^r &\leq \frac{n^m}{m!} e^{mr/n} \nonumber \\
&= \frac{(n e^{r/n})^m} {m!} \nonumber \\
&\rightarrow \lambda ^m / m! \nonumber
\end{align}

@@

As $n \rightarrow \infty$, $\left(1-\frac{m}{n}\right)^m \rightarrow 1$ and $1/m!$ is constant.
We use the Taylor series of $\log(1-t)$ for $0 \leq t \leq 1/2$
to bound the rest.
So
\begin{align}
\log(1-t) &= 0 - t - t^2/2 -t^3/3 - ... \nonumber\\
&\geq -t -t^2/2(1 + 1/2 + 1/4 + ...) = -t - t^2 \nonumber
\end{align}
and we have
\begin{align}
\log \left(n^m\left(1-\frac{m}{n}\right)^r \right)
&= m \log n + r \log\left(1-\frac{m}{n}\right) \nonumber \\
& \geq m \log n - r m /n - r m/n^2. \nonumber
\end{align}

\todo{why?}
Our assumption that $ne^{r/n} \rightarrow \lambda$ implies that
\begin{align}
r = n \log n - n \log \lambda + o(n) \nonumber
\end{align}
so $r(m/n)^2 \rightarrow 0$.
Then
\begin{align}
rm/n &=m \log n - m \log \lambda + o(n)m/n \nonumber \\
m \log \lambda &\rightarrow m \log n - rm/n \nonumber
\end{align}
and
\begin{align}
\log \left(n^m\left(1-\frac{m}{n}\right)^r \right)
&\geq m \log n - rm/n \nonumber
\end{align}
so
\begin{align}
\lim_{n\rightarrow \infty} n^m\left(1-\frac{m}{n}\right)^r
&\geq \lambda^m. \nonumber
\end{align}

Now we piece together the pieces.
To begin,
\begin{align}
{ n \choose m } (1-\frac{m}{n})^r \rightarrow \lambda^m /m! \nonumber
\end{align}
so
\begin{align}
p_0(r,n) \rightarrow \sum_{k=0}^\infty \lambda^k /k! = e^{-\lambda}. \nonumber
\end{align}

The result holds
\begin{align}
p_0(r,n-m) \rightarrow e^{-\lambda} \nonumber
\end{align}
for fixed $m$ since $(n-m)e^{r/(n-m)} \rightarrow \lambda$.

@@

Set $r = n \log n + nx$.
Then 
\begin{align}
ne^{-r/n} &= n e ^{- \log n - x} \nonumber \\
&= n e {- \log n} e^ {- x} = e^{-x}\nonumber
\end{align}

Consider the cover time variable $\cov$.
Since $\cov \leq m$ if and only if the first $m$ steps
of the random walk visit all $n$ vertices, then by \cref{thm:pois},
\begin{align}
\pr(\cov \leq r) &\rightarrow e^{-e^{-x}} \nonumber \\
\pr(\cov - n \log n\leq nx) &\rightarrow e^{-e^{-x}} \nonumber
\end{align}

>>>>

\subsection{Matthews Method on Binary Trees}

\begin{figure}[ht]
	\centering
	\begin{tikzpicture}[scale = 2]
	    \tikzstyle{vertex}=[draw, circle,fill=black, scale=.7]
	    \tikzstyle{edge}=[draw,line width = .5pt,-,black!100]
	    \node[vertex] (0) at (0,3) {};
	    \node[vertex] (1) at (1,2) {};
	    \node[vertex] (2) at (-1,2) {};
	    \node[vertex] (3) at (1.5,1) {};
	    \node[vertex] (4) at (.5,1) {};
	    \node[vertex] (5) at (-.5,1) {};
	    \node[vertex] (6) at (-1.5,1) {};
	    \node[vertex] (7) at (1.75,0) {};
	    \node[vertex] (8) at (1.25,0) {};
	    \node[vertex] (9) at (.75,0) {};
	    \node[vertex] (10) at (.25,0) {};
	    \node[vertex] (11) at (-.25,0) {};
	    \node[vertex] (12) at (-.75,0) {};
	    \node[vertex] (13) at (-1.25,0) {};
	    \node[vertex] (14) at (-1.75,0) {};
	    \draw[edge] (1) -- (0);
	    \draw[edge] (2) -- (0);
	    \draw[edge] (1) -- (3);
	    \draw[edge] (4) -- (1);
	    \draw[edge] (2) -- (5);
	    \draw[edge] (6) -- (2);
	    \draw[edge] (3) -- (7);
	    \draw[edge] (3) -- (8);
	    \draw[edge] (4) -- (9);
	    \draw[edge] (4) -- (10);
	    \draw[edge] (5) -- (11);
	    \draw[edge] (5) -- (12);
	    \draw[edge] (6) -- (13);
	    \draw[edge] (6) -- (14);
	\end{tikzpicture}
	\caption{A binary tree with $k=3$ and $n=15$.}\label{fig:binary}
\end{figure}

\cite{LP17}
Consider an unweighted binary tree as in \cref{fig:binary} with a single root $\rho$.
A vertex has two children if it is distance less than $k$
from the root and no children if it is a leaf (distance $k$ from the root).
There are $n = 1 + 2 + 4 + ... + 2^k = 2^{k+1} - 1$ vertices.

@@

\begin{figure}[ht]
	\centering
	\begin{tikzpicture}[scale = 2]
	    \tikzstyle{vertex}=[draw, circle,fill=black, scale=.7]
	    \tikzstyle{edge}=[draw,line width = .25pt,-,black!100]
	    \tikzstyle{span}=[draw,line width = 2pt,-,black!100]
	    \node[vertex] (0) at (0,0) {};
	    \node[vertex] (1) at (1,0) {};
	    \node[vertex] (2) at (1,1) {};
	    \node[vertex] (3) at (0,1) {};
	    \draw[span] (0) -- (1);
	    \draw[edge] (0) -- (2);
	    \draw[edge] (0) -- (3);
	    \draw[span] (1) -- (2);
	    \draw[edge] (1) -- (3);
	    \draw[span] (2) -- (3);
	\end{tikzpicture}
	\caption{A spanning tree of the complete graph with $n=4$.}\label{fig:span}
\end{figure}

A spanning tree $\T$ is a subset of a graph with all edges
in $V(G)$ connected by the minimum number of edges in $E(G)$.
\cref{fig:span} is an example where the bold edges indicate the edges in $E(\T)$
and the light edges indicate the edges in $E(G)$ but not $E(\T)$.
For the connected graphs we consider, the minimum number of edges
needed to connect all vertices is $n-1$.
Formally, $V(\T) = V(G)$ and $E(\T) \subseteq E(G)$
such that $|E(\T)| = n-1$ and $\T$ is connected.

@@

The expected time of this particular path on the original graph is certainly less than
or equal to the cover time. That is,
\begin{align}
\tcov &\leq \sum_{i=0}^{2n-3} \E_{v_i}[T_{v_{i+1}}]. \nonumber
\end{align}

@@

Using \cref{eqn:commute}, we have $E_x[T_y] + \E_y[T_x] = C \Rxy$ and,
\begin{align}
\tcov &\leq C \sum_{x,y \in E(\T)} \Rxy. \nonumber
\end{align}

@@

If the graph is unweighted, then $C_{x,y} = 1$ and $C = 2 |E(G)|$.
Then
\begin{align}
\tcov \leq 2 |E(G)| (n-1). \nonumber
\end{align}

@@

\section{Distributional Aspects}
\cite{Du11}
We consider the distribution of the cover time of a random walk
on the complete graph with self edges.
Call $r$ the number of steps since we began the random walk.
Let $A_i$ be the event that vertex $i$ is not visited and
$N_n$ the number of unvisited vertices.
Then
\begin{align}
P(A_i) = (1-1/n)^r \nonumber
\end{align}
and
\begin{align}
\E[N_n] = n(1-1/n)^r. \nonumber
\end{align}

If $r/n \rightarrow c$, then
\begin{align}
\E[N_n]/n &= (1-1/n)^r = [(1-1/n)^n]^{r/n} \rightarrow (e^{-1})^c
\rightarrow e^{-c}. \nonumber
\end{align}

We compute the variance of $N_n$ by finding $\E[N_n^2]$ and $\E[N_n]^2$.
Let $\I_m$ be the indicator random variable for event $A_m$.
That is, $\I_m = 1$ if vertex $m$ is unvisited 0 otherwise.
We already know $\E[N_n]$ so we want to find $\E[N_n^2]$,
\begin{align}
\E[N_n^2] &= \E \left[ \left(\sum_{m=1}^n \I_m \right) ^2 \right] =
\E \left[ \sum_{m=1}^n \sum_{k=1}^n \I_m \I_k \right] =
\sum_{1 \leq k, m \leq n} \pr(A_k \cap A_m). \nonumber
\end{align}

The second equality holds since we distribute terms.
The third equality holds since the mean of an indicator for event $A_m$
is the probability of $A_m$.
Putting both terms together, we find the variance
\begin{align}
\Var[N_n] &= \E[N_n^2] - \E[N_n]^2 \nonumber \\
&= \sum_{1 \leq k, m \leq n} \pr(A_k \cap A_m) - P(A_k)P(A_m) \nonumber \\
&= n(n-1) \{ \pr(A_k \cap A_m) - P(A_k)P(A_m) \} + 
n \{ \pr(A_k \cap A_m) - P(A_k)P(A_m) \} \nonumber \\
&= n(n-1) \{ (1-2/n)^r - (1-1/n)^{2r} \} + 
n \{ (1-1/n)^r  - (1-1/n)^{2r} \}. \nonumber
\end{align}
where $k\neq m$ in the first term and $k=m$ in the second.

@@

Thus we conclude that $N_n/n \rightarrow \E[N_n/n] = e^{-c}$ in probability.

The Poisson approximation of the binomial implies that if $n \rightarrow \infty$
and $r/n \rightarrow c$, then the number of visits to each vertex will approach
a Poisson distribution with mean $c$.
We can use this to intuite our conclusion: the proportion of visits to each vertex
approaches $e^{-c}$.

We prove the following theorem and then use it to show that $\cov \sim \textrm{Pois}(e^{-x})$.
\begin{theorem}\label{thm:pois}
If $ne^{-r/n} \rightarrow \lambda \in [0, \infty)$, then the number of unvisited
vertices approaches a Poisson distribution with mean $\lambda$.
\end{theorem}

\begin{proof}
Observe that the probability that there are $k$ unvisited vertices is
\begin{align}
(1-k/n)^r \nonumber.
\end{align}

@@

Our strategy is to show that $p_m(r,n) \rightarrow \frac{\lambda^m}{m!}e^{-\lambda}$.
The first step is to prove that
\begin{align}
{n \choose m} \left(1-\frac{m}{n}\right)^r \rightarrow \lambda ^m /m! \nonumber
\end{align}

We bound the expression from the top and the bottom and use the dominated convergence
theorem.
We will first show the upper bound.
Notice that $(1-x) \leq e^{-x}$ and recall our assumption that
$ne^{r/n} \rightarrow \lambda$.
Then
\begin{align}
{n \choose m} \left(1-\frac{m}{n}\right)^r &\leq \frac{n^m}{m!} e^{mr/n} \nonumber \\
&= \frac{(n e^{r/n})^m} {m!} \nonumber \\
&\rightarrow \frac{\lambda ^m} {m!} \nonumber
\end{align}

@@

As $n \rightarrow \infty$, $\left(1-\frac{m}{n}\right)^m \rightarrow 1$
and $1/m!$ is constant.
We use the Taylor series of $\log(1-t)$ for $0 \leq t \leq 1/2$
to bound the rest of the term.
Notice that
\begin{align}
\log(1-t) &= 0 - t - t^2/2 -t^3/3 - ... \nonumber\\
&\geq -t -t^2/2(1 + 1/2 + 1/4 + ...) = -t - t^2 \nonumber
\end{align}
and so we have
\begin{align}
\log \left(n^m\left(1-\frac{m}{n}\right)^r \right)
&= m \log n + r \log\left(1-\frac{m}{n}\right) \nonumber \\
& \geq m \log n - r m /n - r m/n^2. \nonumber
\end{align}

\todo{why?}
Our assumption that $ne^{r/n} \rightarrow \lambda$ implies that
\begin{align}
r = n \log n - n \log \lambda + o(n) \nonumber
\end{align}
so $r(m/n)^2 \rightarrow 0$.
Then
\begin{align}
rm/n &=m \log n - m \log \lambda + o(n)m/n \nonumber
\end{align}
and in particular
\begin{align}
m \log \lambda &\rightarrow m \log n - rm/n \nonumber
\end{align}
so
\begin{align}
\log \left(n^m\left(1-\frac{m}{n}\right)^r \right)
&\geq m \log n - rm/n. \nonumber
\end{align}
It follows that
\begin{align}
\lim_{n\rightarrow \infty} n^m\left(1-\frac{m}{n}\right)^r
&\geq \lambda^m. \nonumber
\end{align}

Now we put together the pieces.
To begin,
\begin{align}
{ n \choose m } (1-\frac{m}{n})^r \rightarrow \lambda^m /m! \nonumber
\end{align}
so
\begin{align}
p_0(r,n) \rightarrow \sum_{k=0}^\infty \lambda^k /k! = e^{-\lambda}. \nonumber
\end{align}

The result holds for
\begin{align}
p_0(r,n-m) \rightarrow e^{-\lambda} \nonumber
\end{align}
since $m$ is fixed and $(n-m)e^{r/(n-m)} \rightarrow \lambda$.

@@

Set $r = n \log n + nx$.
Then 
\begin{align}
ne^{-r/n} &= n e ^{- \log n - x} \nonumber \\
&= n e {- \log n} e^ {- x} = e^{-x}.\nonumber
\end{align}

Consider the cover time variable $\cov$.
Since $\cov \leq m$ if and only if the first $m$ steps
of the random walk visit all $n$ vertices, then by \cref{thm:pois},
\begin{align}
\pr(\cov \leq r) &\rightarrow e^{-e^{-x}} \nonumber
\end{align}
and
\begin{align}
\pr(\cov - n \log n\leq nx) &\rightarrow e^{-e^{-x}}. \nonumber
\end{align}
